---
title: "klavinde_assignment_5"
output:
  html_document:
    df_print: paged
---

```{r}
library(tidyverse)
library(factoextra)
library(ISLR)
set.seed(123)
cereals.df <- read.csv("cereals.csv")
row.names(cereals.df) <- cereals.df[,1]
cereals.data <- cereals.df[,c(4,5,6,8,9,10,11,12)]
summary(cereals.data)
```

### Using only two variables, Calories and Sugars
```{r}
d <- dist(cereals.data, method = "euclidean")
d.norm <- dist(cereals.data[,c(1,8)], method = "euclidean")
cereals.df.norm <- sapply(cereals.data, scale)
row.names(cereals.df.norm) <- row.names(cereals.data)
d.norm <- dist(cereals.df.norm[,c(1,8)], method = "euclidean")
```

### QUESTION 1. Apply hierarchical clustering to the data using Euclidean distance to the normalized
### measurements. Use Agnes to compare the clustering from single linkage, complete
### linkage, average linkage, and Ward. Choose the best method.
### Compare single linkage, average linkage, complete linkage, and Ward

## a. Using all Variables, Single method
```{r}
d.norm <- dist(cereals.df.norm, method = "euclidean")
hc1 <- hclust(d.norm, method = "single")
memb1 <- cutree(hc1, k = 9)
row.names(cereals.df.norm) <- paste(memb1, ": ", row.names(cereals.data), sep = "")
```

## b. Using all Variables, Average method
```{r}
d.norm <- dist(cereals.df.norm, method = "euclidean")
hc2 <- hclust(d.norm, method = "average")
memb2 <- cutree(hc2, k = 9)
row.names(cereals.df.norm) <- paste(memb2, ": ", row.names(cereals.data), sep = "")
```

## c. Using all Variables, Complete method
```{r}
d.norm <- dist(cereals.df.norm, method = "euclidean")
hc3 <- hclust(d.norm, method = "complete")
memb3 <- cutree(hc3, k = 9)
row.names(cereals.df.norm) <- paste(memb3, ": ", row.names(cereals.data), sep = "")
```

## d. Using all Variables, WARD method
```{r}
d.norm <- dist(cereals.df.norm, method = "euclidean")
hc4 <- hclust(d.norm, method = "ward.D")
memb4 <- cutree(hc4, k = 9)
row.names(cereals.df.norm) <- paste(memb4, ": ", row.names(cereals.data), sep = "")
```

## SINGLE; H-Cluster
```{r}
fviz_dend(x = hc1, cex = 0.4, lwd = 0.7, k = 9,
          k_colors = c("jco"),
          rect = TRUE, 
          rect_border = "jco", 
          rect_fill = TRUE)
```

## AVERAGE; H-Cluster
```{r}
fviz_dend(x = hc2, cex = 0.4, lwd = 0.7, k = 9,
          k_colors = c("jco"),
          rect = TRUE, 
          rect_border = "jco", 
          rect_fill = TRUE) 
```

## COMPLETE; H-Cluster
```{r}
fviz_dend(x = hc3, cex = 0.4, lwd = 0.7, k = 9,
          k_colors = c("jco"),
          rect = TRUE, 
          rect_border = "jco", 
          rect_fill = TRUE)
```

## WARD; H-Cluster
```{r}
fviz_dend(x = hc4, cex = 0.4, lwd = 0.7, k = 9,
          k_colors = c("jco"),
          rect = TRUE, 
          rect_border = "jco", 
          rect_fill = TRUE)
```

### Question 2. 9 Clusters. Using the Ward method and considering 8 variables, stopping at 9 Clusters down would be best. Beyond that doesn't appear meaningful.


## H-Cluster Heat Map of Normalized Cereal Data
```{r}
heatmap(as.matrix(cereals.df.norm), Colv = NA, hclustfun = hclust, 
        col=rev(paste("gray",1:99,sep="")))
```

## Question 4: 
### The Heat Map provides a useful visual for determining most healthy cereals.
### The Labels show the clusters from Ward method. Cluster 1 is high in fiber, low calories, 
### moderate sugar. However the most well-rounded nutritionally is Cluster 5. This cluster 
### is low in fat and sugar, moderate in most everything else (protien, vitamins, fiber)

## Question 3. 
### Partitioned by calories, then summarized to compare 
### averages. Example of isolating cluster for stability not given in 
### "8-Hierarchical Clustering" HTML. Looking forward to the answer 
### release to learn the material. 

```{r}
library(caret)
set.seed(10)
Train_Index = createDataPartition(cereals.data$calories, p=0.6, list = FALSE)
Train_Data = cereals.data[Train_Index, ]
Validation_Data = cereals.data[-Train_Index, ]
TraVal_Data = cereals.data
```
```{r}
summary(Train_Data)
```
```{r}
summary(Validation_Data)
```