---
title: "klavinde_assignment4"
output:
  html_document:
    df_print: paged
---
```{r}
##install.packages("tidyverse")
library(tidyverse) 
##install.packages("factoextra")
library(factoextra)
library(ISLR)
set.seed(123)
pharm.data<-read.csv("Pharmaceuticals.csv")
pharm.trim.data<-pharm.data[,c(3,4,5,6,7,8,9,10,11)]
summary(pharm.trim.data)
df <- scale(pharm.trim.data)
distance <- get_dist(df)
fviz_dist(distance)
fviz_nbclust(df, kmeans, method = "wss")
```
```{r}
fviz_nbclust(df, kmeans, method = "gap_stat")
```
```{r}
fviz_nbclust(df, kmeans, method = "silhouette")
```
## QUESTION a. See Below
##    The WSS elbow is more gradual than the example given in class
##    however it appears that 5 or 6 clusters would be appropriate. 
##    The Silhouette identifies 5 clusters being optimal.
##Using 5 clusters
```{r}
k5 <- kmeans(df, centers = 5, nstart = 21)
k5$centers
```
## QUESTION b. See below

##  Cluster 1 has the highest Price/earnigns. 

##  Cluster 2 grouped the higher estimated revenue growth

##  Cluster 3 appears to be just well diversified, as in a safer portfolio.

##  Cluster 4 is the highest Beta center. 

##  Cluster 5 has the companies with the highest Market Capitalization
```{r}
k5$size
```
##  Cluster 3 also has the highest number of companies (cluster size).
```{r}
fviz_cluster(k5, data = df)
```
## C. Tracing all of the groups back to the .csv table, 
## there does not appear to be a consistent Median Recommendation in the clusters.
## Cluster 3 appears to have the safest diversification of median recommendations.
## Cluster 4 contains companies from each stock exchange (col 12: NSYE, NASDAQ, AMEX), diversifying in that way. 
##d. see below
## Cluster 1: Best Price/Earnings Ratios
## Cluster 2: Best Revenue Growth Companies
## Cluster 3: Most Diversified and Safe Portfolio
## Cluster 4: Best Beta, Diversified Stock Exchanges
## Cluster 5: Best Market Capitalization
##Below gives a visual of the centers for each cluster
```{r}
plot(c(0), xaxt = 'n', ylab = "", type = "l", 
     ylim = c(min(k5$centers), max(k5$centers)), xlim = c(0, 9))
axis(1, at = c(1:9), labels = names(df))
for (i in c(1:9))
  lines(k5$centers[i,1], lty = i, lwd = 2, col = ifelse(i %in% c(0,12),
                                                       "black", "dark grey"))
text(x = 0.5, y = k5$centers[, 1], labels = paste("Cluster", c(1:5)))
```
## Notebook wont put the lines, but they appear on RStudio. Line 66 has an unknown error. 
