---
title: "klavinde_final_exam"
output:
  html_document:
    df_print: paged
---



# Cleaning Data is a large portion of initial coding.

```{r}
arpack <- c("arules","arulesViz")
# install.packages(arpack) install if needed
library(arules)
library(arulesViz)
library(tidyverse)
library(cluster)
library(caret)
fuel_data.df <- read.csv("fuel_receipts_costs.csv")
summary(fuel_data.df)
```



```{r}
row.names(fuel_data.df) <- fuel_data.df[,1]
fuel_numeric<-fuel_data.df%>%column_to_rownames(var = "rowid")
summary(fuel_numeric)
```

## There are many categorical data points and some columns have a large amount of "NA" data points.

```{r}
library(tidyverse)
library(cluster)
summary(fuel_data.df)
colSums(is.na(fuel_data.df))
```



## Fuel Type converted to numerical data to include in clustering
```{r}
fuel_numeric ["fuel_group_code"][fuel_numeric ["fuel_group_code"] == ''] <- NA
fuel_numeric <-fuel_numeric [-which(is.na(fuel_numeric$fuel_group_code)), ]
summary(fuel_numeric)
fuel_numeric$fuel_group_code <- as.numeric(as.factor(fuel_numeric$fuel_group_code))
summary(fuel_numeric)
```

## Fuel Types 1. Coal | 2. Natural Gas | 3. Other Gas | 4. Petroleum | 5. Petroleum Coke
### Alphabetized

```{r}
fuel_numeric <- fuel_numeric[,c(10,14,15,16,17,18,19)]
summary(fuel_numeric)
```
## Data Cleaned! To numerical and without N/A data points.
```{r}
fuel_numeric<- na.omit(fuel_numeric) %>% select_if(is.numeric)
summary(fuel_numeric)
```
# Final Cleaning


### Sampled data down to ~3000 observations using partitioning.
```{r}
library(caret)
set.seed(4381)
Fuel_Index = createDataPartition(fuel_numeric$fuel_mmbtu_per_unit, p=0.988, list = FALSE)
Fuel_Valid = fuel_numeric[Fuel_Index, ]
Fuel_Train = fuel_numeric[-Fuel_Index, ]
head(Fuel_Train)
scaled_fuel_data <- as.data.frame(scale(Fuel_Train))
fuel_euclidean <- dist(scaled_fuel_data, method = "euclidean")
hc <- hclust(fuel_euclidean, method="ward.D2")
```

```{r}
methods <- c( "average", "single", "complete", "ward") 
names(methods) <- c( "average", "single", "complete", "ward")
ac <- function(x) {
  agnes(scaled_fuel_data, metric = "euclidean", method = x)$ac
}
```


## Ward Method is preferred for this data
```{r}
map_dbl(methods, ac)
```


#### This next step takes the most CPU
```{r}
hc_wards <- agnes(scaled_fuel_data, method = "ward", metric = "euclidean")
```


```{r}
hc_wards
library(factoextra)
hc_wards_cut <- cutree(hc_wards, k=7)
table(hc_wards_cut)
pltree(hc_wards, cex = 0.6, hang = -1)
rect.hclust(hc_wards, k = 7, border = 1:5)
fviz_dend(hc_wards, cex = 0.6, k = 7)
```
## First Division: Right is Natural Gas, left is Coal, Petroleum, and Petroleum Coke.

```{r}
library(caret)
library(factoextra)
fuel_data_clustered <- mutate(scaled_fuel_data, cluster = hc_wards_cut)
set.seed(2457)
index <- createDataPartition(fuel_data_clustered$cluster, p = 0.7, list = FALSE)
part_A <- fuel_data_clustered[index,]
part_B <- fuel_data_clustered[-index,]
```

```{r}
hc_wards_A <- agnes(part_A[,1:7], method = "ward", metric = "euclidean")
```

```{r}
hc_cut_A <- cutree(hc_wards_A, k=7)
part_A <- mutate(part_A, cluster = hc_cut_A)
part_A_centroids <- aggregate(part_A[,1:7], by=list(part_A$cluster), mean) 
colnames(part_A_centroids)[1] <- "cluster"
cluster_prediction_B <- data.frame(data=seq(1,nrow(part_B)), Partition_B_cluster=rep(0,nrow(part_B)))
euclidean <- function(a, b) sqrt(sum((a - b)^2))
```

```{r}
for (i in 1:nrow(part_B)) {
  distances <- apply(part_A_centroids[, -1], 1, function(x) euclidean(x, part_B[i,1:7]))
  cluster_prediction_B$Partition_B_cluster[i] <- which.min(distances)
}
```

```{r}
cluster_prediction_B <- cluster_prediction_B %>% mutate(original_clusters = part_B$cluster)
```

```{r}
table(cluster_prediction_B$Partition_B_cluster, cluster_prediction_B$original_clusters, dnn = c("New Cluster", "Old Cluster"))
```
# HEAT MAP of cleaned, normalized data. 
```{r}
heatmap(as.matrix(scaled_fuel_data), Colv = NA, hclustfun = hclust)
```
# This portion is the most useful for reports. 
```{r}
cluster <- hc_wards_cut
centers <- aggregate( . ~ cluster, data = Fuel_Train, FUN = mean)
centers
```

# Cluster 1: Dirtier Coal
### Predominate Fuel Type: Coal
### Quantity (1.8 % of data) 
### Thermal Content (Very High) 
### Sulfur Content: Moderate
### Ash Content: High
### Mercury Content: High
### Fuel Cost Per thermal units $2.31

# Cluster 2: Slightly Better Natural Gas
### Predominate Fuel Type: Natural Gas
### Quantity (4.59 % of data) 
### Thermal Content (Low) 
### Sulfur Content: None/Scant 
### Ash Content: None/Scant 
### Mercury Content: None/Scant 
### Fuel Cost Per thermal units $7.49

# Cluster 3: Petroleum 
### Predominate Fuel Type: Petroleum 
### Quantity (0.2 % of data) 
### Thermal Content (Moderate) 
### Sulfur Content: Low  
### Ash Content: None/Scant 
### Mercury Content: None/Scant 
### Fuel Cost Per thermal units $16.78

# Cluster 4: Standard Natural Gas
### Predominate Fuel Type: Natural Gas 
### Quantity (90.33 % of data)  
### Thermal Content (Low) 
### Sulfur Content: None/Scant  
### Ash Content: None/Scant 
### Mercury Content: None/Scant 
### Fuel Cost Per thermal units $3.62

# Cluster 5: Low Mercury Coal
### Predominate Fuel Type: Coal 
### Quantity (1.77 % of data) 
### Thermal Content (Very High) 
### Sulfur Content: Moderate  
### Ash Content: Higher 
### Mercury Content: Low
### Fuel Cost Per thermal units $2.45

# Cluster 6: High-Ash Coal
### Predominate Fuel Type: Coal (mixed)
### Quantity (1.27 % of data) 
### Thermal Content (Very High) 
### Sulfur Content: Very High
### Ash Content: Highest 
### Mercury Content: None/Scant 
### Fuel Cost Per thermal units $2.37

# Cluster 7: 
### Predominate Fuel Type: Natural Gas
### Quantity (0.013 % of data) | 
### Thermal Content (Low, higher than other Natural Gas) 
### Sulfur Content: None/Scant 
### Ash Content: None/Scant 
### Mercury Content: None/Scant  
### Fuel Cost Per thermal units $31,052 (?? Ultra clean ??)



# Below are Various Charts 

## Chart of all the variables
## Quantity of units overpowers other variables
```{r}
all_fuel <- centers[, c(1:8)] %>% pivot_longer(col = !cluster, names_to = "variable", values_to = "value")

ggplot(data = all_fuel, aes(x = variable, y = value, fill = cluster)) + geom_col(colour = "black") + ggtitle("All Fuel")
```



## FUEL GROUP (TYPE)
### Normalizing data makes fuel group chart difficult to interpret, use center data above for type of fuel. 
### Size of bar height is indicator of type (shorter is coal, tallest is petroleum). 
### Categorical data converted for clustering. 
```{r}
fuel.group.plot <- centers[, c(1,2)] %>% pivot_longer(col = !cluster, names_to = "variable", values_to = "value")

ggplot(data = fuel.group.plot, aes(x = variable, y = value, fill = cluster)) + geom_col(colour = "black") + ggtitle("Fuel Group")
```

## FUEL RECEIVED UNITS
### Visual of how Standard Natural Gas predominates sample
```{r}
fuel.received.units <- centers[, c(1,3)] %>% pivot_longer(col = !cluster, names_to = "variable", values_to = "value")

ggplot(data = fuel.received.units, aes(x = variable, y = value, fill = cluster)) + geom_col(colour = "black") + ggtitle("Fuel Received Units")
```

## FUEL MMBTU PER UNIT
### Clusters 1, 5 , 6 make the most heat per unit (all coal)
```{r}
fuel.mmbtu.per.unit <- centers[, c(1,4)] %>% pivot_longer(col = !cluster, names_to = "variable", values_to = "value")
ggplot(data = fuel.mmbtu.per.unit, aes(x = variable, y = value, fill = cluster)) + geom_col(colour = "black") + ggtitle("fuel mmbtu per unit")
```

## Sulfur Content 
### The clusters with scant to no sulfur are hard to see on bar chart.
```{r}
sulfur.content.pct <- centers[, c(1,5)] %>% pivot_longer(col = !cluster, names_to = "variable", values_to = "value")

ggplot(data = sulfur.content.pct, aes(x = variable, y = value, fill = cluster)) + geom_col(colour = "black") + ggtitle("Sulfur Content")
```



## Ash Content
### Same as Sulfur, the natural gas clusters aren't visible. 
```{r}
ash.content <- centers[, c(1,6)] %>% pivot_longer(col = !cluster, names_to = "variable", values_to = "value")

ggplot(data = ash.content , aes(x = variable, y = value, fill = cluster)) + geom_col(colour = "black") + ggtitle("Ash Content")
```

## Mercury Content
### Predominately one cluster
```{r}
mercury.content <- centers[, c(1,7)] %>% pivot_longer(col = !cluster, names_to = "variable", values_to = "value")

ggplot(data = mercury.content , aes(x = variable, y = value, fill = cluster)) + geom_col(colour = "black") + ggtitle("Mercury Content")
```

## Fuel Cost Per MMBTU
### One type of natural gas was extremely expensive and skewed data
### use center data above for costs.
### Suspect there may be a special purpose natural gas that must be ultra clean.
```{r}
fuel.cost.per.mmbtu <- centers[, c(1,8)] %>% pivot_longer(col = !cluster, names_to = "variable", values_to = "value")

ggplot(data = fuel.cost.per.mmbtu, aes(x = variable, y = value, fill = cluster)) + geom_col(colour = "black") + ggtitle("Fuel Cost Per MMBTU")
```
## To improve report
### Possible further conversion of categorical data (transportation type).
### Ensure Cost per unit is complete and accurate.
### Eliminate blank and N/A in important columns. If none/Scant, put zero. 
